---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
targetFilename <- file.path(".", "activity.csv")
data <- read.csv(targetFilename, header=TRUE, sep=",", stringsAsFactors=FALSE, na.strings="NA", colClasses=c("numeric", "Date", "numeric"))
```

## What is mean total number of steps taken per day?
```{r}
stepsPerDay <- aggregate(steps ~ date, data=data, FUN=sum)
hist(stepsPerDay$steps, breaks=12)
meanStepsPerDay <- mean(stepsPerDay$steps)
medianStepsPerDay <- median(stepsPerDay$steps)
```

## What is the average daily activity pattern?
```{r}
meanStepsPerInterval <- aggregate(steps ~ interval, data=data, FUN=mean)
plot(x=meanStepsPerInterval$interval, y=meanStepsPerInterval$steps, type="l")
meanStepsPerInterval <- aggregate(steps ~ interval, data=data, FUN=mean)
rowWithHighestAverage <- meanStepsPerInterval[meanStepsPerInterval$steps==max(meanStepsPerInterval$steps), ]
intervalWithHighestAverage <- rowWithHighestAverage$interval
abline(v=intervalWithHighestAverage, col="red")
```
  
The interval with the largest mean number of steps is the `r intervalWithHighestAverage`th interval, as marked by the red vertical line in the plot.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
